A Simple Type Theory for Category Theory

### 2022-04-23 paper/paper.tex
sketch of the GAT

@@

To simplify our initiality proof, we use an \emph{essentially
algebraic} presentation of our notion of model. To describe this, we
give a ``combinator'' syntax for VE hyperdoctrines, where we have
explicit sorts for contexts $\Gamma$ and $\Phi$ and explicit judgments
for substitutions $\gamma : \Delta \to \Gamma$ and $\phi : \Psi
\vdash_{f,g} \Phi$.
%
While the explicit description of the theory is quite long, we can
describe it in simple modular pieces.
%
First, we use the standard theory of a category with families, which
describes the algebraic structure of contexts $\Gamma$, types $\Gamma
\vdash A$ and terms $\Gamma \vdash M : A$.
%
Second, we add to that a theory for a slight generalization of virtual
double categories we call \emph{double categories with display
pro-arrows}.
%
The difference is that in a double category with display pro-arrows,
we have a true double category with strict vertical and horizontal
composition, among whom are the ``display'' pro-arrows.
%
The display pro-arrows model the profunctors of the syntax, which
don't have a strict composition, whereas the arbitrary pro-arrows
model the contexts $\Phi$, which do have a strict composition, given
by context concatenation $\Phi,\Psi$.
%
The difference is that in a virtual double category the contexts are
assumed to be given inductively by formal compositions of the display
pro-arrows, whereas in models of this algebraic theory, the contexts
might include other exotic objects.
%
This difference is analogous to the difference between a category with
families and a contextual category in models of dependent type theory,
and we fully expect that there is a similar relationship between the
models: virtual double categories should form a co-reflective
subcategory characterized as those double categories with display
pro-arrows where all pro-arrows are equal to some concatenation of
displays.
%
Finally, we parameterize all of the constructions of the theory of
double categories with pro-arrows by contexts $\Gamma$ and add in
equations that substitution commutes with all double category
operations.
%
This means the theory presents a category with families with an double
category with display pro-arrows internal to the category of
presheaves on the category of contexts, i.e., a hyperdoctrine of
double categories with display pro-arrows.

@@

\begin{figure}
  \begin{mathpar}
    \inferrule
    {}
    {\Gamma : \isaCtx}

    \inferrule
    {\Delta : \isaCtx \and \Gamma : \isaCtx}
    {\gamma : \textrm{Subst}(\Delta,\Gamma)}

    \inferrule
    {\Gamma : \isaCtx}
    {C : \textrm{Cat}(\Gamma)}

    \inferrule
    {\Gamma : \isaCtx \and C : \textrm{Cat}(\Gamma) \and D : \textrm{Cat}(\Gamma) \and
    }
    {F : \textrm{Functor}(\Gamma,C,D)}

    \inferrule
    {\Gamma : \isaCtx \and C : \textrm{Cat}(\Gamma) \and D : \textrm{Cat}(\Gamma) \and
    }
    {P : \textrm{Profunctor}(\Gamma,C,D)}

    \inferrule
    {\Gamma : \isaCtx \and C : \textrm{Cat}(\Gamma) \and D : \textrm{Cat}(\Gamma)
    }
    {P : \textrm{ProfCtx}(\Gamma,C,D)}

    \inferrule
    {\Gamma : \textrm{Ctx}\and
      C  : \textrm{Cat}(\Gamma) \and D  : \textrm{Cat}(\Gamma) \and
      C' : \textrm{Cat}(\Gamma) \and D' : \textrm{Cat}(\Gamma) \and
      f : \textrm{Functor}(\Gamma,C,C') \and g : \textrm{Cat}(\Gamma,D,D') \and
      \Phi : \textrm{ProfCtx}(\Gamma,C,D)\and
      \Psi : \textrm{ProfCtx}(\Gamma,C',D')
    }
    {\phi : \textrm{ProfSubst}(\Gamma,C,D,C',D',f,g,\Phi,\Psi)}

    \inferrule
    {\Gamma : \textrm{Ctx}\and
      C  : \textrm{Cat}(\Gamma) \and D  : \textrm{Cat}(\Gamma) \and
      C' : \textrm{Cat}(\Gamma) \and D' : \textrm{Cat}(\Gamma) \and
      f : \textrm{Functor}(\Gamma,C,C') \and g : \textrm{Cat}(\Gamma,D,D') \and
      \Phi : \textrm{ProfCtx}(\Gamma,C,D)\and
      P : \textrm{ProfCtx}(\Gamma,C',D')
    }
    {t : \textrm{Trans}(\Gamma,C,D,C',D',f,g,\Phi,P)}
  \end{mathpar}
  \caption{Sorts of Algebraic SCT}
\end{figure}

@@

\begin{figure}
  \begin{mathpar}
    \inferrule
    {g : \textrm{Functor}(D,E) \and \textrm{Functor}(C,D)}
    {g \circ_{\Gamma,C,D,E} f : \textrm{Functor}(C,E)}

    (h \circ g) \circ f = h \circ (g \circ f)

    \inferrule
    {}
    {\id_C : \textrm{Functor}(C,C)}

    (h \circ \id) = h

    \id \circ h = h

    \inferrule
    {P : \textrm{Profunctor}(\Gamma,C',D')
    \and f : \textrm{Functor}(\Gamma,C,C')\and
    \and g : \textrm{Functor}(\Gamma,D,D')}
    {P \circ_{\Gamma,C,D,C',D'} (f,g) : \textrm{Profunctor}(\Gamma,C,D)}

    P \circ (f,g) \circ (f',g') = P \circ (f \circ f', g \circ g')

    P \circ (\id,\id) = P

    \inferrule
    {\psi : \textrm{ProfSubst}(f',g',\Phi,\Psi)\and \phi : \textrm{ProfSubst}(f,g,\Sigma,\Phi)
    }
    {\psi \circ \phi : \textrm{ProfSubst}(f' \circ f, g' \circ g, \Sigma, \Psi)}

    \inferrule
    {}
    {restrict(f,g,\phi) : \textrm{ProfSubst}(\id,\id,\Psi,\Phi)}
  \end{mathpar}
  \caption{Algebraic Presentation of SCT}
\end{figure}

@@

%% \begin{definition}
%%   For any $C : \textrm{Cat}$, we can define an identity functor
%%   $\textrm{Id}_C : \textrm{Functor} C C$ by
%%   \begin{enumerate}
%%   \item $\textrm{Id}_0 = \lambda \alpha:C_0. \alpha$
%%   \item $\textrm{Id}_1 = \pendlambdaXdotY \alpha
%%     {\homrlambdaXatYdotZ f {\alpha'} f}$
%%   \item Preservation of identity requires proving
%%     \[
%%     \lambda \alpha. \pendappXtoY {C_r}{\textrm{Id}_0\alpha} \]
%%     is equal to
%%     \[ \lambda \alpha.
%%     \homrappXtoYatZ {\pendappXtoY {\textrm{Id}_1} \alpha} {(\pendappXtoY {C_r}\alpha)} \alpha\]
%%     Both sides simply $\beta$ reduce to $ \lambda \alpha. \pendappXtoY {C_r} \alpha$
%%     and the result follows by reflexivity.
%%   \item Preservation of composition requires proving (under a $\lambda \alpha. \homrlambdaXatYdotZ f \beta \homrlambdaXatYdotZ g \gamma$)
%%     \[
%%     \homrappXtoYatZ {\pendappXtoY {{\textrm{Id}}_1} \alpha} {(\homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {C_t} \alpha} f \beta} g \gamma)}\gamma
%%     = \homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {C_t} {{\textrm{Id}}_0 \alpha}} {(\homrappXtoYatZ {\pendappXtoY {{\textrm{Id}}_1} \alpha} f \beta)} {{\textrm{Id}}_0 \beta}} {(\homrappXtoYatZ {\pendappXtoY{{\textrm{Id}}_1} \beta} g \gamma)} {{\textrm{Id}}_0 \gamma}\]
%%     Both sides $\beta$ reduce to

@@

%% \begin{definition}
%%   Given categories $C,D,E$ and functors $F : \textrm{Functor} C D$ and
%%   $G : \textrm{Functor} D E$, we define the composition $G \circ F :
%%   \textrm{Functor} C E$ as
%%   \begin{enumerate}
%%   \item $(G \circ F)_0 = \lambda \alpha. {G_0 ({F_0} \alpha)}$
%%   \item $(G \circ F)_1 =
%%     \pendlambdaXdotY \alpha {
%%     \homrlambdaXatYdotZ f \beta
%%     {\homrappXtoYatZ {\pendappXtoY {G_1} {F_0 \alpha}} {\homrappXtoYatZ {\pendappXtoY {F_1} \alpha} f \beta} {F_0 \beta}}}$
%%   \item Preservation of identity requires proving
%%     \[ \lambda \alpha. \pendappXtoY {E_r} {(G \circ F)_0\alpha} \]
%%     is equal to
%%     \[ \lambda \alpha.
%%     \homrappXtoYatZ {\pendappXtoY {(G \circ F)_1} \alpha} {\pendappXtoY {C_r}\alpha} \alpha \]

@@

%% First, we have the ``algebraic'' notion of a functor representing a
%% profunctor, i.e., when a profunctor $R : C\pto D$ is equivalent to
%% $Hom_D(-,G=)$.
%% \begin{definition}[Right Representability]
%%   Let $C, D : \text{Cat}$ and $Q : \textrm{Prof}\,C D$ be a profunctor
%%   between them. Then a \emph{right representation} of $Q$ consists of
%%   \begin{enumerate}
%%   \item A functor $G : \textrm{Functor}\, D C$
%%   \item An isomorphism $I,I^-$ between $Q$ and $\text{Hom}(C)[Id;G]$
%%   \end{enumerate}
%% \end{definition}

>>>>

% Algebraic notation
\newcommand{\algCtx}{\textrm{Ctx}}
\newcommand{\algSubst}{\textrm{Subst}}
\newcommand{\algTy}{\textrm{Type}}
\newcommand{\algTm}{\textrm{Term}}
\newcommand{\algCat}{\textrm{Cat}}
\newcommand{\algVarr}{\textrm{Functor}}
\newcommand{\algHCtx}{\textrm{ProfCtx}}
\newcommand{\algHSubst}{\textrm{ProfSubst}}
\newcommand{\algHarr}{\textrm{Prof}}
\newcommand{\algTrans}{\textrm{Trans}}

@@

To simplify our initiality proof, we use a \emph{generalized
algebraic} presentation of our notion of
model\cite{cartmell,sterling-eat}. To describe this, we give a
``combinator'' syntax for VE hyperdoctrines, where we have explicit
sorts for contexts $\Gamma$ and $\Phi$ and explicit judgments for
substitutions $\gamma : \Delta \to \Gamma$ and $\phi : \Psi
\vdash_{f,g} \Phi$.
%
While the explicit description of the theory is quite long, we can
describe it in simple modular pieces.
%
First, we use the standard theory of a category with families, which
describes the algebraic structure of contexts $\Gamma$, types $\Gamma
\vdash A$ and terms $\Gamma \vdash M : A$.
%
Second, we add to that a theory for a slight generalization of virtual
double categories we call \emph{double categories with display
pro-arrows}.
%
The difference is that in a double category with display pro-arrows,
we have a true double category with strict vertical and horizontal
composition, among whom are the ``display'' pro-arrows.
%
The display pro-arrows model the profunctors of the syntax, which
don't have a strict composition, whereas the arbitrary pro-arrows
model the contexts $\Phi$, which do have a strict composition, given
by context concatenation $\Phi,\Psi$.
%
The difference is that in a virtual double category the contexts are
assumed to be given inductively by formal compositions of the display
pro-arrows, whereas in models of this algebraic theory, the contexts
might include other exotic objects.
%
This difference is analogous to the difference between a category with
families and a contextual category in models of dependent type theory,
and we fully expect that there is a similar relationship between the
models: virtual double categories should form a co-reflective
subcategory characterized as those double categories with display
pro-arrows where all pro-arrows are equal to some concatenation of
displays.
%
Finally, we parameterize all of the constructions of the theory of
double categories with pro-arrows by contexts $\Gamma$ and add in
equations that substitution commutes with all double category
operations.
%
This means the theory presents a category with families with an double
category with display pro-arrows internal to the category of
presheaves on the category of contexts, i.e., a hyperdoctrine of
double categories with display pro-arrows.

@@

\begin{figure}
  \begin{mathpar}
    \inferrule
    {}
    {\Gamma : \algCtx}

    \inferrule
    {\Delta : \algCtx \and \Gamma : \algCtx}
    {\gamma : \algSubst(\Delta,\Gamma)}

    \inferrule
    {\Gamma : \algCtx}
    {C : \algCat(\Gamma)}

    \inferrule
    {\Gamma : \algCtx \and C : \algCat(\Gamma) \and D : \algCat(\Gamma) \and
    }
    {F : \algVarr(\Gamma,C,D)}

    \inferrule
    {\Gamma : \algCtx \and C : \algCat(\Gamma) \and D : \algCat(\Gamma) \and
    }
    {P : \textrm{Profunctor}(\Gamma,C,D)}

    \inferrule
    {\Gamma : \algCtx \and C : \algCat(\Gamma) \and D : \algCat(\Gamma)
    }
    {P : \textrm{ProfCtx}(\Gamma,C,D)}

    \inferrule
    {\Gamma : \textrm{Ctx}\and
      C  : \algCat(\Gamma) \and D  : \algCat(\Gamma) \and
      C' : \algCat(\Gamma) \and D' : \algCat(\Gamma) \and
      f : \algVarr(\Gamma,C,C') \and g : \algCat(\Gamma,D,D') \and
      \Phi : \textrm{ProfCtx}(\Gamma,C,D)\and
      \Psi : \textrm{ProfCtx}(\Gamma,C',D')
    }
    {\phi : \textrm{ProfSubst}(\Gamma,C,D,C',D',f,g,\Phi,\Psi)}

    \inferrule
    {\Gamma : \textrm{Ctx}\and
      C  : \algCat(\Gamma) \and D  : \algCat(\Gamma) \and
      C' : \algCat(\Gamma) \and D' : \algCat(\Gamma) \and
      f : \algVarr(\Gamma,C,C') \and g : \algCat(\Gamma,D,D') \and
      \Phi : \textrm{ProfCtx}(\Gamma,C,D)\and
      P : \textrm{ProfCtx}(\Gamma,C',D')
    }
    {t : \textrm{Trans}(\Gamma,C,D,C',D',f,g,\Phi,P)}
  \end{mathpar}
  \caption{Sorts of Algebraic SCT}
\end{figure}

@@

The algebraic theory of a double category with displays indexed by a
category is given in Figure~\ref{fig:alg-dc-disp}.
%
First, we give the categories and functors the theory of a category
indexed by a category.
%
Next, we give categories and \emph{profunctor contexts} the same
theory, where the identity $\id_h$ is the ``empty'' context $\alpha:C$ and the composition $\Phi \circ_h$.
%

\begin{figure}
  \begin{mathpar}
    % theory of an indexed category (vertical)
    \inferrule
    {g : \algVarr(D,E) \and \algVarr(C,D)}
    {g \circ_v f : \algVarr(C,E)}

    (g \circ_v f) \circ_{vs} \gamma = (g \circ_{vs} \gamma) \circ_v (f \circ_s \gamma)

    (h \circ_v g) \circ_v f = h \circ_v (g \circ_v f)

    \inferrule
    {}
    {\id_h : \algVarr(C,C)}

    \id_h \circ_{vs} \gamma = \id_h

    (h \circ \id) = h\and \id \circ h = h

    % theory of an indexed category (horizontal)

    % embedding of displays

    % strict equipment (horizontal)
    % strict equipment  (displays)

    % 2-cells: substitutions
    % vertical category
    % horizontal category
    % interchange law

    % 2-cells: transformations : Q: is any of this necessary or should
    % we just conflate transformations with substitutions into a
    % display?
    % 
    % action of trans on subst with assoc, unit
    % identity trans with unit
    % embedding of transformations into substitutions

    \inferrule
    {P : \algHarr(\Gamma,C',D')
    \and f : \algVarr(\Gamma,C,C')\and
    \and g : \algVarr(\Gamma,D,D')}
    {P \circ_r (f,g) : \algHarr(\Gamma,C,D)}

    (P \circ_r (f,g) \circ_s \gamma) = (P \circ_s \gamma) \circ_r (f \circ_s \gamma, g \circ_s \gamma)

    P \circ_r (f,g) \circ_r (f',g') = P \circ_r (f \circ_r f', g \circ_r g')

    P \circ_r (\id,\id) = P

    \inferrule
    {\psi : \algHSubst(f',g',\Phi,\Psi)\and \phi : \algHSubst(f,g,\Sigma,\Phi)
    }
    {\psi \circ \phi : \algHSubst(f' \circ f, g' \circ g, \Sigma, \Psi)}

    \inferrule
    {}
    {restrict(f,g,\phi) : \algHSubst(\id,\id,\Psi,\Phi)}
  \end{mathpar}
  \caption{Equipment with Displays}
  \label{fig:alg-dc-disp}
\end{figure}

\begin{figure}
  \begin{mathpar}
    % Cat Multiplicatives
    % Cat of Elements
    % Presheaf cats

    % Prof Multiplicatives
    % Unit
    % Tensor
    % Hom

    % Internalization
    % Cats
    % Functor
    % Prof
    % ProfCtx?
    % Trans

    % Cat Additives
    % Product
    % Terminal

    % Prof Additives
    % Product
    % Terminal
  \end{mathpar}
  \caption{UMPs in an Equipment}
\end{figure}

@@

%% \begin{definition}
%%   For any $C : \algCat$, we can define an identity functor
%%   $\textrm{Id}_C : \algVarr C C$ by
%%   \begin{enumerate}
%%   \item $\textrm{Id}_0 = \lambda \alpha:C_0. \alpha$
%%   \item $\textrm{Id}_1 = \pendlambdaXdotY \alpha
%%     {\homrlambdaXatYdotZ f {\alpha'} f}$
%%   \item Preservation of identity requires proving
%%     \[
%%     \lambda \alpha. \pendappXtoY {C_r}{\textrm{Id}_0\alpha} \]
%%     is equal to
%%     \[ \lambda \alpha.
%%     \homrappXtoYatZ {\pendappXtoY {\textrm{Id}_1} \alpha} {(\pendappXtoY {C_r}\alpha)} \alpha\]
%%     Both sides simply $\beta$ reduce to $ \lambda \alpha. \pendappXtoY {C_r} \alpha$
%%     and the result follows by reflexivity.
%%   \item Preservation of composition requires proving (under a $\lambda \alpha. \homrlambdaXatYdotZ f \beta \homrlambdaXatYdotZ g \gamma$)
%%     \[
%%     \homrappXtoYatZ {\pendappXtoY {{\textrm{Id}}_1} \alpha} {(\homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {C_t} \alpha} f \beta} g \gamma)}\gamma
%%     = \homrappXtoYatZ {\homrappXtoYatZ {\pendappXtoY {C_t} {{\textrm{Id}}_0 \alpha}} {(\homrappXtoYatZ {\pendappXtoY {{\textrm{Id}}_1} \alpha} f \beta)} {{\textrm{Id}}_0 \beta}} {(\homrappXtoYatZ {\pendappXtoY{{\textrm{Id}}_1} \beta} g \gamma)} {{\textrm{Id}}_0 \gamma}\]
%%     Both sides $\beta$ reduce to

@@

%% \begin{definition}
%%   Given categories $C,D,E$ and functors $F : \algVarr C D$ and
%%   $G : \algVarr D E$, we define the composition $G \circ F :
%%   \algVarr C E$ as
%%   \begin{enumerate}
%%   \item $(G \circ F)_0 = \lambda \alpha. {G_0 ({F_0} \alpha)}$
%%   \item $(G \circ F)_1 =
%%     \pendlambdaXdotY \alpha {
%%     \homrlambdaXatYdotZ f \beta
%%     {\homrappXtoYatZ {\pendappXtoY {G_1} {F_0 \alpha}} {\homrappXtoYatZ {\pendappXtoY {F_1} \alpha} f \beta} {F_0 \beta}}}$
%%   \item Preservation of identity requires proving
%%     \[ \lambda \alpha. \pendappXtoY {E_r} {(G \circ F)_0\alpha} \]
%%     is equal to
%%     \[ \lambda \alpha.
%%     \homrappXtoYatZ {\pendappXtoY {(G \circ F)_1} \alpha} {\pendappXtoY {C_r}\alpha} \alpha \]

@@

%% First, we have the ``algebraic'' notion of a functor representing a
%% profunctor, i.e., when a profunctor $R : C\pto D$ is equivalent to
%% $Hom_D(-,G=)$.
%% \begin{definition}[Right Representability]
%%   Let $C, D : \text{Cat}$ and $Q : \textrm{Prof}\,C D$ be a profunctor
%%   between them. Then a \emph{right representation} of $Q$ consists of
%%   \begin{enumerate}
%%   \item A functor $G : \algVarr\, D C$
%%   \item An isomorphism $I,I^-$ between $Q$ and $\text{Hom}(C)[Id;G]$
%%   \end{enumerate}
%% \end{definition}
